Investigating soundscapes perception through acoustic scenes simulation

### 2018-09-03 lafaySimulatedScenePerception.tex
GL relecture

%Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches. Il permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed here can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment. With such material, not only is the type of sound sources available for study, but also the exact level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}. Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}. Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written that study how humans discriminate different kind of urban soundscape, see~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} .

@@

@@

%La littérature fait généralement la distinction entre :
% \begin{itemize}
% \item {un événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
% \item {une texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps \cite{saint1995classification}.
% \end{itemize}

@@

% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.

%Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.

%Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée \cite{bregman1994auditory}. Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}. Inversement, quand le cerveau ne parvient pas à isoler des événements, les éléments constituants de la scène sont agglomérés dans un même flux.

%According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}. Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.

%Maffiolo~\cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores. Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».

%Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}). Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores. Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»). Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux. %Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans~\cite{guastavino2006ideal,dubois2006cognitive}.

%Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.

%Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.

%Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.

% mcdermott2013summary

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Figure~\ref{fig:taxonomieEb}. Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.

@@

%La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé. Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.

%En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection. Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this can be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions. For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.

@@

@@

44 students (30 male, 14 female; averaging $21.6$ years of age, s.d. of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part in the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.

@@

Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).

@@

All the subjects performed the experiment on computers with equivalent hardware and software configurations. The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli were the scenes obtained in experiment 1.a. The output sound level was the same for all the subjects.

@@

All the subjects performed the experiment simultaneously. They were not allowed to talk to each other during the experiment.

@@

An experimenter was available during the whole duration of the experiment in order to assist subjects and answer any questions the subjects might have.

@@

10 students (8 male, 2 female; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} took part to the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing. None of them took part to the previous simulation experiment (experiment 1.a).

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
                                   &                            \\
                                   &                           \\
                                   &                          \\
                                   &                     \\
                                   &                    \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the experiments.}
\label{tab:acronyme}
\end{table}

@@

A set of features, upon which the analysis is conducted, is attached to each sound scene. A summary of those features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes.

@@

\item \emph{structural features}: while \emph{SimScene} allows us to acces a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels. To figure those out, we draw inspiration from the $L_{Aeq}$. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non-ideal scenes during the experiment 1.a. To do so, we consider if there exists significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non-ideal scenes. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

%\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène a quoi sert Asubject ici ?} \gl{Les deux sont possibles : Ascene mesure des différences structurelles, Asubject des différences perceptives}

% GL comprends pas en francais

%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels, nous évaluons cette significativité à partir d'un test de Student à deux populations;

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non-ideal scenes. The significance is measured with a two-sample Student test.

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness. To do so, we consider linear correlations between this feature and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$. Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$. Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$. Les classes typiques sont nommées \textbf{marqueurs sonores};

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$. Such typical classes are called \textbf{sound markers}, in reference to the work of \cite{schafer1993soundscape}. Testing is done for each class, at each level of abstraction, and separately for texture and event classes.

@@

\item \emph{Representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene. Denoting as $S_i$ the semantic features of scene $i$, we compute the distances between all $S_i$ vectors. A Hamming distance is used: considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors. It is defined as follows:

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes.

%\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif. + plaisante que les ni-scene. ca revient au même. c'est les test qui les écarte. il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.} \gl{Oui. Mais ça revient au même de toute façon}

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

However, level alone is not sufficient to fully differentiate between ideal and non-ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & ideal scenes          & non-ideal scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not bring much information, whatever the type of environment is.

@@

%\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?} \gl{Elle n'est pas parlante}

@@

To conclude, the level of non-ideal markers has a negative influence on the pleasantness for the non-ideal scenes. On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.

@@

\subsubsection*{Experimental protocol}

>>>>

%Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches. Elle permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).

@@

For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed here can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment. With such material, not only is the type of sound sources available for study, but also the exact sound level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}. Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}. Conclusions and discussion about future work follow in Section~\ref{sec:conclusion}.

@@

The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts that study how humans discriminate different kind of urban soundscape, see~\cite{leobon_analyse_1986,maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,beaumont2004pertinence,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,polack2008perceptive,niessen2010categories,brown2011towards} .

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for sound events and one for sound textures. Up to four levels of abstraction are considered from the most generic classes(level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Appendix~\ref{app:taxonomie}. Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé. Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique. En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection. Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this may be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions. For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

44 students (30 male, 14 female; averaging $21.6$ years of age, s.d. of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) performed the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.

@@

The subjects evaluate the 80 simulated scenes generated in experiment 1.a. Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes (from sec. 15 to sec. 45).

@@

All the subjects performed the experiment on standard desktop computers with the same hardware and software configurations. The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli were the scenes obtained in experiment 1.a. The output sound level was the same for all the subjects.

@@

All the subjects performed the experiment simultaneously in a quiet environment. They were not allowed to talk to each other during the experiment.

@@

An experimenter was available during the whole duration of the experiment in order to assist subjects and to answer questions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

10 students (8 male, 2 female; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} performed the experiment. All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing. None of them took part in the previous simulation experiment (experiment 1.a).

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the experiments.}
\label{tab:acronyme}
\end{table}

@@

A set of features, upon which the analysis is conducted, is attached to each sound scene. A summary of those features (and the corresponding acronyms) is presented in Table~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for each scene, computed as the average of all the scores given by all the subjects to a specific scene. $\mathcal{A}_{subject}$ is the average pleasantness for each subject, computed as the average of all the scores given by a specific subject to all the scenes. $\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes separately. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.

@@

\item \emph{structural features}: while \emph{SimScene} allows us to access a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels. To figure those out, we draw inspiration from the $L_{Aeq}$. In our case, it is a scalar computed from the signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during the experiment 1.a. To do so, we consider if there exists significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels. Nous évaluons cette significativité à partir d'un test de Student à deux populations;

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels ($L$, $L(E)$ and $L(T)$) differ between the ideal and non-ideal scenes. The significance is measured with a two-sample Student test.

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels ($L$, $L(E)$ and $L(T)$) impact the perceived pleasantness. To do so, we consider linear correlations between those features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$. Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used in a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of time a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

% Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$. Les classes typiques sont nommées \textbf{marqueurs sonores};

If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of environnement $k$. Such typical classes are called \textbf{sound markers}, in reference to the work of \cite{schafer1993soundscape}. Testing is done for each class, at each level of abstraction, and separately for texture and event classes.

@@

\item \emph{Representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scenes. Denoting as $S_i$ the semantic features of scene $i$, we compute the distances between all $S_i$ vectors. A Hamming distance is used: considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors. It is defined as follows:

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p<0.01$) than the non-ideal scenes.

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$, mean deviation: -7 $dB$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

However, sound level alone is not sufficient to fully differentiate between ideal and non-ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & ideal scenes          & non-ideal scenes      \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$) \\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$) \\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$)         \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes distinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$ remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($r=-0.32$, $p=0.06$) for ideal scenes.

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes ($r=-0.53$, $p<0.01$) disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not bring much information, whatever the type of environment is.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

To conclude, the level of non-ideal markers has a negative influence on the pleasantness for the non-ideal scenes. On the other hand, the level of ideal markers does not influence the perceived pleasantness for the ideal scenes.

@@
